\section{Comparing ArtDev3D to CGP}
We've now seen that implementations of the models using the framework, were able to reproduce the results achieved in experiments conducted by their inventors. However, there is one other purpose of this framework that we still need to discuss. That is the ability to compare two models with a minimum set of interfering factors.


\subsection{Design}
The design of CGP is very elegant. It is based on a very simple system of a digital circuit. The cell does not need to do much other than generating input and parsing output. In contrast, ArtDev3D handles much more complex, lower level activities such as protein synthesis and protein requests. It is deeper rooted in biology, a property that is reflected in the original code. ArtDev3D also has an ``easier'' cell program. It is easier to see what it does, because most of it is in plain sight. CGP does not have this quality as the whole cell program is evolved. This makes it harder to analyze and see what it really does. We do know that it uses chemicals to a great extent. It is information that is exchanged between all cells. This is something that is not implemented in ArtDev3D. In fact, that is why chemicals have been reported\cite{hoye2006} to complicate things. Further investigation can determine whether or not chemicals really are useful in general, or whether exchanging cell type information is sufficient.


\subsection{Stability}
From the results we've seen so far, there are a number of things that can be said about these models. Most prominently, we see that ArtDev3D is able to evolve perfect specimens within relatively short time. With the sphere experiment, perfect fitness was achieved within just 500 generations. What can't be seen from the results is that it would achieve this typically within 50 generations. In contrast, the CGP model requires a lot of generations in order to reach higher fitness. At 10 000 generations, fitness rarely exceeded 0.9. Even when doubling that number, the population was still improving. Obviously, one cannot set it to run forever. The problem seemed to lie in the development method used in the models.

In ArtDev3D, the development or growth would stop after a certain amount of steps regardless of how long the development time. This ensures that the organisms will grow to target shape without overgrowing. H{\o}ye\cite{hoye2006} demonstrated this property in his thesis, and this is reflected by the model's ability to achieve high fitness fast. From the results obtained, it seemed that it is difficult to correctly evolve desired phenotype without knowing how many development steps will be needed. For instance, with the French flag experiment, using four development steps was optimal compared to three or five. Three was obviously not enough because the cells do not have the time to grow properly, while five was too much because the cells started to ``deform''. The new test showed this tendency as well, but it is difficult to state anything for certain with these results. As I've mentioned, one will have to develop a tool to look into each development step ascertain this behaviour.
